A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-02-14 jacobian_calc_CMAME.tex
added reason for changes as notes

\note[Michael]{re: I.2}\change[Michael]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics}. The related mathematical
definitions and parts of peridynamic theory especially relevant to the present
work are explained in the sequel.

@@

\subsubsection{Vector states for state-based peridynamic theory.}

@@

\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simualtion.}
The tangent stiffness matrix is the frechet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
equilibrium version of the peridynamic equation of motion. The internal force
density is computed in the peridynamic equation of motion by computing the
volume integral over the whole body of the Frechet derivative of the 
strain energy state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

For the finite-difference type methods in \emph{Peridigm}, the perturbation method is employed
by probing the deformed configuration one degree of freedom at a time and so allowing the
sensitivity of the force density state wrt the deformation state for all nodes to be computed.

\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++ STL vectors,
are hard coded to be of type {\tt double} and can not be simply be declared as
having a complex data type. Therefore, program methods that were used in the
course of applying the FD method were copied, renamed and re-written to follow
the CS formula. This involved changing the data type of intermediate variables
which were locally scoped, and dynamically casting persistent memory variables
to complex data types where necessary. No attempt was made to ``tune'' the CS
code in any way for performance. The algorithm  was implemented in such as way
that it was in one-to-one correspondence with the FD and CD methods; however,
it must be noted that the overhead associated with the dynamic casts in the CS
method likely hurts its overall performance and this issue could be alleviated
by using the next generation of templated \emph{TPetra} vectors available in
\emph{Trilinos} that are capable of taking an explicit instantiation of any
data type including {\tt complex}. The version of \emph{Peridigm} modified to
implement CS can be found in the paper repository.

>>>>

\note[Michael]{re: I.2, John I think you better handle expanding this section,
I was going to essentially copy silling:psa.}\change[Michael]{Briefly, peridynamics}Peridynamics \cite{silling2000ret,silling:psa,silling2010peridynamic}
is a nonlocal reformulation of the partial-differential equations that provide
the statement of momentum balance in classical continuum mechanics. Its primary
goal is to avoid the use of spatial derivatives in the balance equations or
constitutive laws such that discontinuous displacements, i.e. cracks, are
mathematically consistent with the governing equations. It has demonstrated
great promise in modeling problems with pervasive failure
\cite{littlewood2010}, interesting phenomenon such as dynamic crack branching
\cite{ha2010sod}, and is being extended to multiphyics phenomena
\cite{bobaru2011peridynamic,katiyar2013} and the problem of multi-scale
coupling to molecular simulations \cite{seleson2009peridynamics}. The related mathematical
definitions and parts of peridynamic theory especially relevant to the present
work are explained in the sequel.

@@

\note[Michael]{re: I.2. John, this was to be the skeleton for the new sections on peridynamics.}
\subsubsection{Vector states for state-based peridynamic theory.}

@@

\note[Michael]{re: I.2} \add[Michael]{
\subsubsection{The tangent stiffness matrix in a quasi-static peridynamic simualtion.}
The tangent stiffness matrix is the frechet derivative of the residual wrt the
deformation state.  For the purposes of this work, the residual represents the unbalanced forces in the
equilibrium version of the peridynamic equation of motion. The internal force
density is computed in the peridynamic equation of motion by computing the
volume integral over, a horizon, of the Frechet derivative of the 
strain energy state wrt to the deformed configuration inner product with the deformation state, per node,
for each of the three directions in the original reference frame.

For the finite-difference type methods in \emph{Peridigm}, the perturbation method is employed
by probing the deformed configuration one degree of freedom at a time and so allowing the
sensitivity of the force density state wrt the deformation state for all nodes to be computed.
}
\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows
closely the implementation of the FD technique.  However, because of
\emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the
\emph{Epetra} package, special steps had to be followed to allow the use of
complex number data types.  This is because \emph{Epetra} vectors, which can be
thought of as distributed memory parallel cousins to standard C++ STL vectors,
are hard coded to be of type {\tt double} and can not be simply be declared as
having a complex data type. Therefore, program methods that were used in the
course of applying the FD method were copied, renamed and re-written to follow
the CS formula. This involved changing the data type of intermediate variables
which were locally scoped, and dynamically casting persistent memory variables
to complex data types where necessary. No attempt was made to ``tune'' the CS
code in any way for performance. The algorithm  was implemented in such as way
that it was in one-to-one correspondence with the FD and CD methods; however,
it must be noted that the overhead associated with the dynamic casts in the CS
method likely hurts its overall performance and this issue could be alleviated
by using the next generation of templated \emph{TPetra} vectors available in
\emph{Trilinos} that are capable of taking an explicit instantiation of any
data type including {\tt complex}. The version of \emph{Peridigm} modified to
implement CS can be found in the paper repository.
